Formalizing Stack Safety as a Security Property

### 2023-02-04 submission5/core.tex
squeeze on IX; didn't help as much as I hoped

We are not attempting to present a universal definition of stack safety
for all systems. While many security properties can be described at
the level of a high-level programming language and translated to a
target machine by a secure compiler, stack safety cannot be defined in
this way, since ``the stack'' is completely transparent in the
definitions of most source languages.\footnote{Contrast Azevedo de
  Amorim et al.'s work on heap safety
  \cite{DBLP:conf/post/AmorimHP18}: the concept of the heap figures
  directly in high-level language semantics and its security is
  therefore amenable to a high-level treatment.}
%
But neither is it satisfying as an entirely low-level property;
indeed, at the lowest level, the specification of a ``well-behaved
stack'' is almost vacuous. The ISA is not concerned with such
questions as whether a caller's frame should be readable or writable
to its callee. Those are the purview of high-level languages built
atop the hardware stack.

Therefore, any low-level treatment of stack safety must begin by asking:
which high-level features does our system support, and what are their
security requirements? Different answers will beget different properties.
We identify and define five security requirements that a system might try
to enforce upon a function call: well-bracketed control flow, and the
integrity and confidentiality of each of the caller and the callee.
These are by no means exhaustive. We apply these first to a simple system
with very few features, then to a more realistic one supporting
tailcall-elimination, argument passing on the stack, and callee-saves
registers.

We show that our properties are useful for distinguishing correct enforcement
from incorrect using QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based
testing tool for Coq. We generate random programs, using the features of our
more realistic system, and check that the stack safety {\em micro-policies}
of Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}
correctly detect the ones that attempt to violate one of our properties.

@@

\sna{See commented out tex for content that could be massaged to fit in here,
  depending on whether we're interested in delving deeper.}
%For instance, the stack-safety specification from Georges et al.
%\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
%control flow, and therefore defines {\em Well-bracketed Control Flow} (\(\wbcf\)) under the assumption that
%a callee should always return to its caller. But, in the presence of tail-call elimination, a
%return from a callee to a distant ancestor is perfectly reasonable. Similarly, the preceding work by
%Skorstengaard et al. \cite{SkorstengaardSTKJFP} made no mention of confidentiality, even though
%their enforcement mechanism did protect the confidentiality of the caller (though not the callee).

%In particular, many enforcement techniques focus purely on
%well-bracketed control flow. For instance, stack canaries aim to prevent certain attacks on the return
%address, and shadow stacks with protection (e.g. Return Address Defender~\cite{Chiueh2001RAD})
%to enforce it completely. Others combine this protection with some degree of memory protection,
%chiefly focusing on integrity. Interestingly, Skorstengaard et al.~\cite{SkorstengaardSTKJFP}
%describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
%to the combination of \(\clri\) and \(\clrc\). In fact, the follow-up work by
%Georges et al. \cite{Georges22:TempsDesCerises} was inspired by the properties presented in this paper to extend their formal discussion to include confidentiality.
%When checking if a preliminary version of these properties applied to
%the Cerise calling convention, they noted that the Skostengaard et al. calling convention
%did not enforce callee confidentiality, and made sure that their version would
%\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
%This demonstrates the benefit of our choice to explicitly state properties in security
%terms: specifying security mechanisms is hard, and when the specification takes the
%form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
%security requirement.

@@

\subsection{Testing The Stack Safety Menagerie}

In \cref{sec:intro}, we discussed several dimensions of flexibility
that our properties should exhibit: applicability to different system
features and different security requirements. Another important dimension
is applicability to multiple enforcement mechanisms.
In the following, we discuss how different existing mechanisms
might fit into our model.

\paragraph{Stack Canaries and Shadow Stacks}
%
Canaries are special values inserted by the caller at certain points of the
stack and checked for corruption on return, for protection against
some classes of control-flow hijacking (i.e., overwriting the caller's
return address through a buffer overflow that would also overwrite the canary
detectably). A shadow stack is a mirror image of the control-relevant parts of the stack,
maintained by the program to attempt to detect attempts to hijack its control
flow and ``restrict the flexibility available in creating gadget chains''
\cite{Shanbhogue+19}.
%
Interestingly, these are lazy enforcement mechanisms, in that
the attack may occur and be detected some time later, as long as
it is detected before it can become dangerous. That would make our
formalism a good fit for defining their security, except that
they are chiefly hardening techniques: they increase the difficulty
of some control-flow attacks on the stack, but cannot provide absolute
guarantees on \(\wbcf\) under a normal attacker model.

\paragraph{Split Stacks}
%
Code-pointer integrity seeks to prevent control-flow hijacking attacks by
splitting the program memory, including the stack, into a regular region and a
safe region; objects that require protection are placed in the latter and
all accesses to them protected by static or dynamic checks.
%
Like shadow stacks, bounds checking is a high-level technique designed to
protect code written in a high-level language, but it is designed to
offer protection under a stronger threat model that also considers the presence
of untrusted code. As with most other software-based methods, it purports to
protect exclusively the control-relevant parts of the stack, making no claims
about its data-relevant parts---we would solely be testing \(\wbcf\).

\paragraph{Bounds Checking}
%
Under a bounds checking discipline such as SoftBound \cite{NagarakatteZMZ09}, all the pointers
in a program are extended with some disjoint metadata, and these are combined
to gate memory accesses. These approaches enforce a form of \emph{memory safety},
and we would therefore expect them to enforce \(\clri\) and \(\clrc\). They aim
to enforce \(\wbcf\) by cutting off attacks that involve memory-safety violations,
but that may not be sufficient.

Bounds checking approaches require substantial compiler cooperation. This is not a
problem for our properties in general, but it is not very compatible with
generation-by-execution of low-level code. A better choice would be to generate
high-level code using a tool like CSmith \cite{DBLP:conf/pldi/YangCER11}, or prove the properties instead.

\paragraph{Capabilities}

Capability machines can represent and manipulate
unforgeable tokens of authority over specific regions of
memory using a combination of dedicated out-of-band memory and
specialized ISA instructions.
%
The most prominent contemporary representative of this long line of work is
CHERI \cite{DBLP:conf/sp/WatsonWNMACDDGL15}, a modern architecture designed to provide efficient fine-grained
memory protection and compartmentalization.
%
Previous work has used these principles in simplified models to implement
\emph{secure calling conventions} for the call
stack \cite{SkorstengaardLocal,SkorstengaardSTKJFP,Georges22:TempsDesCerises}. Those conventions can be seen as combining the
protections of bounds-checking with software-based methods of control-flow protection.

There are several proposals around the use of CHERI capabilities to enforce stack safety,
including mechanisms that use the standard CHERI hardware (which includes local
capabilities) \cite{SkorstengaardLocal},
and others that propose entirely new types of capabilities, such as linear
\cite{SkorstengaardSTK}, uninitialized \cite{Georges+21}, lifetime
\cite{Tsampas+19}, and directed \cite{Georges22:TempsDesCerises} capabilities.
It is a high priority to test this most recent work by Georges et al., as it is designed
to be similar to all of our properties except for \(\clei\).

\subsection{More Features}

Several common features did not make it into this work: most importantly concurrency and
exceptions. We plan to extend the model to include them, and then test them. We would also
like to test the model given in \cref{app:ptr} for arbitrary memory-safe pointer sharing.

\subsection{Proofs}

>>>>

\apt{Too much real estate to sell the generality, flexibility, etc. of our approach.
  Again, we are not providing evidence that these attributes are true, although we can
  and should conjecture claim them in passing.}

\bcp{Agreed.  I think in particular the first paragraph can just go
  (I've deleted it).
  But this is not enough.  Also, the section is trying to do at least
  two things---talk about earlier work and how we compare, and talk
  about why we believe a flexible approach is good.  The whole
  discussion is a bit wooly and confusing.}

\sna{Should probably move the discussion of the CHERI work to related work,
  and focus on justifying our approach.}
\bcp{+1}

% Security properties should be motivated by security requirements
% and flexible enough to support various sets of security requirements depending on the
% goals of the system that they validate.
% Following this reasoning, our vision of stack safety is as a collection of properties
% describing the security needs of a system with regard to its functional abstractions.
% While the stack still plays a key role, our stack safety is about functions. The most
% dramatic consequence of this is that we treat registers as part of a function activation's
% state, and hold them to the same protection requirements as stack memory.

Flexibility is important for talking rigorously about stack safety.
%
The reason is that, while many security properties can be described at
the level of a high-level programming language and translated to a
target machine by a secure compiler, stack safety cannot be defined in
this way, since ``the stack'' is completely transparent in the
definitions of most source languages.\footnote{Contrast Azevedo de
  Amorim et al.'s work on heap safety
  \cite{DBLP:conf/post/AmorimHP18}: the concept of the heap figures
  directly in high-level language semantics and its security is
  therefore amenable to a high-level treatment.}
%
But neither is it satisfying as an entirely low-level property;
indeed, at the lowest level, the specification of a ``well-behaved
stack'' is almost vacuous. The ISA is not concerned with such
questions as whether a caller's frame should be readable or writable
to its callee: those are the purview of high-level languages built
atop the hardware stack.

\bcp{There's a broken transition here: the paragraph below is not an
  instance of the paragraph above.  Indeed, I don't see how they
  connect at all.}

For instance, the stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\em Well-bracketed Control Flow} (\(\wbcf\)) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Similarly, the preceding work by
Skorstengaard et al. \cite{SkorstengaardSTKJFP} made no mention of confidentiality, even though
their enforcement mechanism did protect the confidentiality of the caller (though not the callee).

These two cases represent two dimensions in which a stack safety definition requires
flexibility: what constitutes safe behavior differs between systems that support different
features, and different contexts have different top-level security requirements.

We are not attempting to present a universal definition of stack safety for all systems,
but we do construct our definition with these dimensions of flexibility in mind.
Our model is decomposed into:
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for the integrity, confidentiality, and control flow
requirements.

We show that we can extend our system to new features by first introducing a simple
example model, and then moving to a more realistic model with multiple new features:
callee-save registers, argument passing on the stack, and tail-call elimination, simply
by refining a few definitions in (1). This becomes the model that we validate.
In \cref{app:ptr}, we extend the model further with arbitrary sharing of memory-safe
pointers. Our criteria in (3) cover a total of five core requirements:
\(\wbcf\), caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
\(\clei\) subsumes \(\clrc\) and \(\clec\) subsumes
\(\clri\), but otherwise the properties are orthogonal and can be taken piecemeal
according to the needs of a system.

In particular, many enforcement techniques focus purely on
well-bracketed control flow. For instance, stack canaries aim to prevent certain attacks on the return
address, and shadow stacks with protection (e.g. Return Address Defender~\cite{Chiueh2001RAD})
to enforce it completely. Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Interestingly, Skorstengaard et al.~\cite{SkorstengaardSTKJFP}
describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
to the combination of \(\clri\) and \(\clrc\). In fact, the follow-up work by
Georges et al. \cite{Georges22:TempsDesCerises} was inspired by the properties presented in this paper to extend their formal discussion to include confidentiality.
When checking if a preliminary version of these properties applied to
the Cerise calling convention, they noted that the Skostengaard et al. calling convention
did not enforce callee confidentiality, and made sure that their version would
\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
This demonstrates the benefit of our choice to explicitly state properties in security
terms: specifying security mechanisms is hard, and when the specification takes the
form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
security requirement.

%These are defined
%formally in \cref{sec:props}. . We further define two
%important combinations of properties: full stack safety \(\textnormal{\sc FSS}\),
%and caller safety \(\textnormal{\sc ClrS}\). Of these, most existing concepts of
%stack safety correspond to either \(\wbcf\) or \(\textnormal{\sc ClrS}\), which are
%concerned with the basic guarantees that the caller recieves during the call, while
%\(\textnormal{\sc FSS}\) is a natural extension that enforces a higher-level concept
%of a correct interface between caller and callee. \sna{Blah. Needs work.}

Flexibility aside, we show that our properties are useful for their purpose:
distinguishing correct enforcement from incorrect. We use
QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based testing
tool for Coq, to generate random programs and check that the
stack safety {\em micro-policies} of Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}
correctly detect the ones that attempt to violate one of our properties.

@@

@@

In \cref{sec:intro}, we discussed several dimensions of flexibility
that our properties should exhibit: applicability to different system
features and different security requirements. Another important dimension
is applicability to multiple enforcement mechanisms. \apt{Could we mention that
  in intro as well?}  There is much future work in each of these dimensions.

% \subsection{Testing The Stack Safety Menagerie}

\paragraph*{Enforcement with capabilities}
Capability machines can represent and manipulate
unforgeable tokens of authority over specific regions of
memory using a combination of dedicated out-of-band memory and
specialized ISA instructions.
%
The most prominent contemporary representative of this long line of work is
CHERI \cite{DBLP:conf/sp/WatsonWNMACDDGL15}, a modern architecture designed to provide efficient fine-grained
memory protection and compartmentalization.
%
Previous work has used these principles in simplified models to implement
\emph{secure calling conventions} for the call
stack \cite{SkorstengaardLocal,SkorstengaardSTKJFP,Georges22:TempsDesCerises}. Those conventions can be seen as combining the
protections of bounds-checking with software-based methods of control-flow protection.
%
%There are several proposals around the use of CHERI capabilities to enforce stack safety,
%including mechanisms that use the standard CHERI hardware (which includes local
%capabilities) \cite{SkorstengaardLocal},
%and others that propose entirely new types of capabilities, such as linear
%\cite{SkorstengaardSTK}, uninitialized \cite{Georges+21}, lifetime
%\cite{Tsampas+19}, and directed \cite{Georges22:TempsDesCerises} capabilities.
It is a high priority to test this most recent work by Georges et al.~\cite{Georges22:TempsDesCerises}, as it is designed
to enforce analogues of all of our properties except for \(\clei\).

\paragraph*{Enforcement in software}
%
Under a bounds checking discipline such as SoftBound \cite{NagarakatteZMZ09}, all the pointers
in a program are extended with some disjoint metadata, and these are combined
to gate memory accesses. These approaches enforce a form of \emph{memory safety},
and we would therefore expect them to enforce \(\clri\) and \(\clrc\). They aim
to enforce \(\wbcf\) by cutting off attacks that involve memory-safety violations,
but that may not be sufficient.
Bounds checking approaches require substantial compiler cooperation. This is not a
problem for our properties in general, but it is not very compatible with
generation-by-execution of low-level code. A better choice might be to generate
high-level code using a tool like CSmith \cite{DBLP:conf/pldi/YangCER11}, or prove the properties instead.

\paragraph*{More Features}

We have preliminary work on extending our model to handle C++-style
exceptions, which, like tailcalls, obey only a weakened version of \(\wbcf\). 
We are also exploring extensions to concurrency, starting with a model of
statically allocated co-routines.  These extensions will also require non-trivial
testing effort.  We also plan to test
the model given in \cref{app:ptr} for arbitrary memory-safe pointer sharing.

\paragraph*{Weaker Requirements}

Several popular enforcement mechanisms are not designed to provide
absolute guarantees of security. For example, stack canaries~\cite{Cowan+98}
and shadow stacks~\cite{Dang+15,Shanbhogue+19}
are chiefly hardening techniques: they increase the difficulty
of some control-flow attacks on the stack, but cannot provide absolute
guarantees on \(\wbcf\) under a normal attacker model.
%
Interestingly, these are lazy enforcement mechanisms, in that
the attack may occur and be detected some time later, as long as
it is detected before it can become dangerous. That would make our
observation-based formalism a good fit for defining their security,
if we could find a formal characterization of what they do acheive
(perhaps in terms of a base machine with restricted addressing power).

%% \paragraph{Split Stacks}
%% %
%% Code-pointer integrity seeks to prevent control-flow hijacking attacks by
%% splitting the program memory, including the stack, into a regular region and a
%% safe region; objects that require protection are placed in the latter and
%% all accesses to them protected by static or dynamic checks.
%% %
%% Like shadow stacks, bounds checking is a high-level technique designed to
%% protect code written in a high-level language, but it is designed to
%% offer protection under a stronger threat model that also considers the presence
%% of untrusted code. As with most other software-based methods, it purports to
%% protect exclusively the control-relevant parts of the stack, making no claims
%% about its data-relevant parts---we would solely be testing \(\wbcf\).

\paragraph*{Proofs}

The present work relies on randomized testing to validate properties versus
enforcement mechanisms, but full formal proofs are certainly possible in principle,
and we would like to investigate whether they are feasible in practice.
Having formalizing our definitions in Coq gives us a head start in this effort.
